Parallel global optimization algorithm employing decision trees for launching local methods

### 2025-10-12 2025/MMST_DecisionTree/ParallelDecisionTree.tex
Добавил расшифровку MPI, замечание первого рецензента  MPI (Message Passing Interface)
reduce similarity

%В настоящей работе рассматриваются параллельные алгоритмы решения многомерных задач глобальной оптимизации. Такие задачи часто возникают в тех случаях, когда необходимо подобрать значения параметров исследуемой математической модели, при которых результаты моделирования лучше всего соответствуют экспериментальным данным. Для решения задач этого класса известно множество алгоритмов: от метаэвристических, основанных на идее случайного поиска \cite{Ferreiro2013,Garcia2014,Langdon2011}, до детерминированных алгоритмов, гарантирующих сходимость к глобальному минимуму \cite{Evtushenko2009,He2008,Paulavicius2011}.
This paper investigates parallel algorithms for solving multidimensional global optimization problems These problems commonly occur when optimizing the parameters of a mathematical model to achieve the best possible agreement between simulation results and experimental observations. Numerous algorithms have been developed to address this class of problems, varying from metaheuristic and evolutionary techniques \cite{Ferreiro2013,Garcia2014,Langdon2011} to deterministic algorithms designed to guarantee convergence to a global minimizer \cite{Evtushenko2009,He2008,Paulavicius2011}.

%Поскольку в реальных задачах глобальной оптимизации каждое вычисление значения функции (далее \textit{trial}) является операцией, требующей больших вычислительных затрат, количество таких попыток приходится сокращать. Это может быть достигнуто за счет преднамеренного выбора вариантов в ходе поиска оптимального решения, отсекая бесперспективные поисковые под области и исследуя только те, в которых можно найти решение задачи. Алгоритм глобального поиска (АГП) основан на этой идее \cite{Strongin2000}. В настоящей работе мы попытались объединить АГП и метод локальной оптимизации \cite{Nocedal, Kelley}  (метод поиска по образцу Хука-Дживса \cite{HookJeeves}), чтобы уменьшить количество выполняемых испытаний. Решение о запуске локального метода будет приниматься с использованием дерева решений.
In real-world global optimization problems, each function evaluation (hereafter referred to as a \textit{trial}) is a computationally expensive operation, making it necessary to reduce the number of such trials. This may be achieved by the purposeful selection of alternatives during the course of optimization, rejecting unpromising search sub-regions and exploring only those where a solution can be found. 
%именно на такой идее основан один из эффективных детерминированных методов глобальной оптимизации - Алгоритм Глобального поиска
This idea forms the foundation of an efficient deterministic global optimization method: the Global Search Algorithm (GSA) \cite{Strongin2000}.
In this work, we attempted to combine the GSA and a local optimization method (the Hooke-Jeeves method \cite{Nocedal, Kelley}) to decrease the number of trials carried out. The decision to run the local method will be made using a decision tree.

@@

We assume that the function satisfies the Lipschitz condition 
\begin{displaymath} 
	|\varphi(y')-\varphi(y'')|\leq L \left\| y'-y'' \right\|, \; y',y'' \in D, 0<L< \infty, 
\end{displaymath}
with a constant $L$ that is unknown a priori.
This condition corresponds to a bounded change in the function's values for a bounded change in the argument.
The assumption that the objective function satisfies the Lipschitz condition can be considered (in the context of applied problems) as a reflection of the bounded possible variations in the system under optimization.

@@

The approach under consideration employs a dimensionality reduction scheme using Peano curves to solve multidimensional problems.
These curves enable the conversion of a multidimensional optimization problem within the region $D$ into a one-dimensional minimization problem over an interval $[0,1]$,
\begin{displaymath}
	\varphi(y(x^\ast))\ =\min\{\varphi(y(x)): x\in [0,1]\},
\end{displaymath}
where function $\varphi(y(x^\ast))$ satisfies the Hölder condition
\begin{displaymath}
	\left|\varphi (y \left(x_1\right))- \varphi (y \left(x_2\right)\right )|\le\ H\left|x_1-x_2\right|^\frac{1}{N},\ x_1,\ x_2\epsilon[0,1].
\end{displaymath}

@@

Fig. \ref{fig:fig2} shows the graph of the function $\varphi(y)$; Fig. \ref{fig:fig2_2} illustrates its approximation $\psi(y)$ constructed using a decision tree.
\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{Graph of the benchmark objective function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

@@

At the initial stage, the master process (let it have the identifier 0) launches $p$ trials in parallel at $p$ different points $\{y\left(x^1\right),y\left(x^2\right),\ldots,y\left(x^p\right)\}$, where preimages $\{x^1,x^2,\ldots,x^p\}$ are from the interval $[0,1]$.
Two preimages are the boundary points $x^1=0, x^p=1$; the remaining preimages are interior points $x^i\in\left(0,1\right),i=2,\ldots,p-1$.
The sets $U$ and $V$ are empty.

@@

	\item  Compute the current lower bound $M$ of the unknown Hölder constant $H$: 
	\begin{equation} 
		\label{agp2_mu} 	\mu=max\left\{\frac{|z_i-z_{i-1}|}{{{(x}_i-x_{i-1})}^{1/N}},\ i=1,\ldots,k\right\},\ M=\  \left\{\begin{matrix}r\mu,\ \mu>0,\\1,\ \mu=0,\\\end{matrix}\right.\ 	
	\end{equation} 
	where $r>1$ is the method parameter. This parameter controls the reliability of the algorithm: higher values of $r$ guarantee finding the global minimum, while choosing a smaller value accelerates the convergence of the algorithm.

	\item  For each interval $(x_{i-1},x_i), 1\leq i\leq k+1,$ compute a value $R(i)$, called \textit{the characteristic} of the interval, according to the following formulas:
	\begin{equation} 
		\label{agp3_R1} R(1)=2\Delta_1-4\dfrac{z_1}{M}, \; R(k+1)=2\Delta_{k+1}-4\dfrac{z_k}{M}, 
	\end{equation} 
	\begin{equation} 
		\label{agp3_Ri} R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{M},1<i<k+1, 
	\end{equation} 
	where \(\Delta_i=(x_i-x_{i-1})^\frac{1}{N}\).

@@

	\item If $k < 100N$, go to Step~1.

	\item Build a decision tree using all the accumulated search information and obtain the approximating function $\psi(y)$.

	\item If the decision tree is being used for the first time, construct a uniform grid
	\begin{displaymath} 
		Y'=\{ y'\in R^N:\ a_i\le y'_i \le b_i,\ 1\le i\le N \},
	\end{displaymath} 
	where the number of grid nodes is a parameter of the method.

	\item Compute the piecewise constant approximation of the function $\psi(y)$ using the decision tree: $Z' = \{ z' = \psi(y'), y' \in Y' \}$.

@@

The next $p$ trials are executed in parallel at the points $x^{k+j},\ 1\leq j\leq p$,  computed using formula  (\ref{agp5_x1}). Upon completion of these trials, their results are stored in the search information database, and the algorithm proceeds to compute new trial points. Note that, generally, the trial process in applied optimization problems is far more computationally expensive than computing the coordinates of a point.

@@

%Численные эксперименты проводились на суперкомпьютере Лобачевский. Каждый узел имел по два процессора Intel Sandy Bridge E5-2660 2,2 ГГц, 64 Гб оперативной памяти.
The Lobachevsky supercomputer was used for the numerical experiments. Each supercomputer node featured two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB of RAM.
%В экспериментах использовался генератор тестовых задач GKLS, который может генерировать задачи многоэкстремальной оптимизации с известными свойствами: точка глобального минимума, количество локальных минимумов и др.
The GKLS test problem generator was used in the experiments. This generator can create multi-extremal optimization problems with known properties, such as the location of the global minimum, the number of local minima, etc.

>>>>

Many decision-making problems arising in various areas of human activity can be formulated as optimization problems. With the rapidly increasing complexity of optimized objects, it is quite natural for their mathematical models to become more complex as well, which consequently complicates the search for an optimal combination of parameters. Very often, it is impossible to find such a combination analytically, making it necessary to construct numerical methods for its search.

%В настоящей работе рассматриваются параллельные алгоритмы решения многомерных задач глобальной оптимизации. Такие задачи часто возникают в тех случаях, когда необходимо подобрать значения параметров исследуемой математической модели, при которых результаты моделирования лучше всего соответствуют экспериментальным данным. Для решения задач этого класса известно множество алгоритмов: от метаэвристических, основанных на идее случайного поиска \cite{Ferreiro2013,Garcia2014,Langdon2011}, до детерминированных алгоритмов, гарантирующих сходимость к глобальному минимуму \cite{Evtushenko2009,He2008,Paulavicius2011}.
This paper investigates parallel algorithms for solving multidimensional global optimization problems. These problems commonly occur when optimizing the parameters of a mathematical model to achieve the best possible agreement between simulation results and experimental observations. Numerous algorithms have been developed to address this class of problems, varying from metaheuristic and evolutionary techniques \cite{Ferreiro2013,Garcia2014,Langdon2011} to deterministic algorithms designed to guarantee convergence to a global minimizer \cite{Evtushenko2009,He2008,Paulavicius2011}.

%Поскольку в реальных задачах глобальной оптимизации каждое вычисление значения функции (далее \textit{trial}) является операцией, требующей больших вычислительных затрат, количество таких попыток приходится сокращать. Это может быть достигнуто за счет преднамеренного выбора вариантов в ходе поиска оптимального решения, отсекая бесперспективные поисковые под области и исследуя только те, в которых можно найти решение задачи. Алгоритм глобального поиска (АГП) основан на этой идее \cite{Strongin2000}. В настоящей работе мы попытались объединить АГП и метод локальной оптимизации \cite{Nocedal, Kelley}  (метод поиска по образцу Хука-Дживса \cite{HookJeeves}), чтобы уменьшить количество выполняемых испытаний. Решение о запуске локального метода будет приниматься с использованием дерева решений.
In real-world global optimization problems, each function evaluation (hereafter referred to as a \textit{trial}) is a computationally expensive operation, making it necessary to reduce the number of such trials. This may be achieved by the purposeful selection of alternatives during the course of optimization, rejecting unpromising search sub-regions and exploring only those where a solution can be found. 
%именно на такой идее основан один из эффективных детерминированных методов глобальной оптимизации - Алгоритм Глобального поиска
This idea forms the foundation of an efficient deterministic global optimization method: the Global Search Algorithm (GSA) \cite{Strongin2000}.
In this work, we attempted to combine the GSA and a local optimization method (the Hooke-Jeeves method \cite{Nocedal, Kelley}) to decrease the number of trials carried out.
A decision tree will determine whether to run the local method.

@@

We assume that the function satisfies the Lipschitz condition 
\begin{displaymath} 
	|\varphi(y_1)-\varphi(y_2)|\leq L \left\| y_1-y_2 \right\|, \; y_1,y_2 \in D, 
\end{displaymath}
with a constant $L, \; 0<L< \infty$,  that is unknown a priori.
This condition corresponds to a bounded change in the function's values for a bounded change in the argument.
The assumption that the objective function satisfies the Lipschitz condition can be considered (in the context of applied problems) as a reflection of the bounded possible variations in the system under optimization.

@@

The approach under consideration employs a dimensionality reduction scheme using Peano curves to solve multidimensional problems.
These curves enable the conversion of a multidimensional optimization problem within the region $D$ into a one-dimensional minimization problem over an interval $[0,1]$,
\begin{displaymath}
	\varphi(y(x^\ast))\ =\min\{\varphi(y(x)): x\in [0,1]\},
\end{displaymath}
where function $\varphi(y(x^\ast))$ satisfies the Hölder condition
\begin{displaymath}
	\left|\varphi (y \left(x_1\right))- \varphi (y \left(x_2\right)\right )|\le\ H\left|x_1-x_2\right|^\frac{1}{N},\ x_1, x_2\in[0,1].
\end{displaymath}

@@

Fig. \ref{fig:fig2} shows the graph of the function $\varphi(y)$; Fig. \ref{fig:fig2_2} illustrates its approximation $\psi(y)$ constructed using a decision tree.
\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{Graph of the benchmark function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

@@

The master process (let it have the identifier 0) initiates $p$ parallel trials, each starting at a uniquely defined point $\{y\left(x^1\right),y\left(x^2\right),\ldots,y\left(x^p\right)\}$, where preimages $\{x^1,x^2,\ldots,x^p\}$ are from the interval $[0,1]$.
Two preimages are the boundary points $x^1=0, x^p=1$; the remaining preimages are interior points $x^i\in\left(0,1\right),i=2,\ldots,p-1$.
The sets $U$ and $V$ are empty.

@@

	\item  Compute the current lower bound $M$ of the unknown Hölder constant $H$: 
	\begin{equation} 
		\label{agp2_mu} 	\mu=max\left\{\frac{|z_i-z_{i-1}|}{{{(x}_i-x_{i-1})}^{1/N}},\ i=1,\ldots,k\right\},\ M=\  \left\{\begin{matrix}r\mu,\ \mu>0,\\1,\ \mu=0,\\\end{matrix}\right.\ 	
	\end{equation} 
	where $r>1$ is the method parameter. Reliability of the algorithm is controlled by this parameter: higher values of $r$ guarantee finding the global minimum, while choosing a smaller value accelerates the convergence of the algorithm.

	\item  
	%For each interval $(x_{i-1},x_i), 1\leq i\leq k+1,$ compute a value $R(i)$, called \textit{the characteristic} of the interval, according to the following formulas:
	For every interval $(x_{i-1},x_i), 1\leq i\leq k+1,$ compute its \textit{characteristic} $R(i)$, as defined by the following formulas:
	\begin{equation} 
		\label{agp3_R1} R(1)=2\Delta_1-4\dfrac{z_1}{M}, \; R(k+1)=2\Delta_{k+1}-4\dfrac{z_k}{M}, 
	\end{equation} 
	\begin{equation} 
		\label{agp3_Ri} R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{M},1<i<k+1, 
	\end{equation} 
	where \(\Delta_i=(x_i-x_{i-1})^\frac{1}{N}\).

@@

	\item If the number of trials is not large enough, namely, if $k < 100 N$, go to Step~1.

	\item Build a decision tree using all the accumulated search information and obtain the approximation $\psi(y)$.

	\item 
	%If the decision tree is being used for the first time, construct a uniform grid
	Upon the first application of the decision tree, generate a uniform grid
	\begin{displaymath} 
		Y'=\{ y'\in R^N:\ a_i\le y'_i \le b_i,\ 1\le i\le N \},
	\end{displaymath} 
	%where the number of grid nodes is a parameter of the method.
	where the density of grid points is governed by a user-defined parameter.

	\item Compute the piecewise-constant approximation $\psi(y)$ of the function $\varphi(y)$ using the decision tree.% $Z' = \{ z' = \psi(y'), y' \in Y' \}$.

@@

The next $p$ trials are carried out in parallel at the points $x^{k+j},\ 1\leq j\leq p$,  computed using formula  (\ref{agp5_x1}). Upon completion of these trials, their results are stored in the search information database, and the algorithm proceeds to compute new trial points. Note that, generally, the trial process in applied optimization problems is far more computationally expensive than computing the coordinates of a point.

@@

%Численные эксперименты проводились на суперкомпьютере Лобачевский. Каждый узел имел по два процессора Intel Sandy Bridge E5-2660 2,2 ГГц, 64 Гб оперативной памяти.
The Lobachevsky supercomputer was used for the numerical experiments. Each supercomputer node featured two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB of RAM.
%В экспериментах использовался генератор тестовых задач GKLS, который может генерировать задачи многоэкстремальной оптимизации с известными свойствами: точка глобального минимума, количество локальных минимумов и др.
The GKLS test problem generator \cite{Sergeyev2006} was used in the experiments.
%This generator can create multi-extremal optimization problems with known properties, such as the location of the global minimum, the number of local minima, etc.
The generator produces complex optimization problems with multiple local minima and predictable characteristics, like the global minimum's location and the number of local minima.
